Using the internal language of toposes in algebraic geometry

### 2015-11-26 slides-ihes2015.tex
Ending the talk with a "funny slide"
Slides almost finished

\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{multicol}
\usepackage{tabto}
\usepackage{xstring}
\usepackage{mathtools}
\usepackage{soul}\setul{0.2ex}{}
\usepackage[all]{xy}
\usepackage{tikz}
\usetikzlibrary{calc,shapes.callouts,shapes.arrows}

@@

@@

\newcommand{\floatbox}[3]{%
  \raisebox{0pt}[0pt][0pt]{%
    \begin{picture}(0,0)(#1,#2)#3\end{picture}\leavevmode%
  }%
}

@@

\note{
  Let~$X$ be a topological space (or locale) and let~$\alpha : \F \to \G$ be a
  morphism of sheaves on~$X$. Then:
  \allowdisplaybreaks
  \begin{align*}
    & \Sh(X) \models \speak{$\alpha$ is injective} \\[0.5em]
    \Longleftrightarrow\
    & \Sh(X) \models \forall s\?\F\_ \forall t\?\F\_ \alpha(s) = \alpha(t) \Rightarrow s = t \\[0.5em]
    \Longleftrightarrow\ &
      \text{for all open~$U \subseteq X$, sections $s \in \F(U)$:} \\
    &\qquad
      \text{for all open~$V \subseteq U$, sections $t \in \F(V)$:} \\
    &\qquad\qquad
        \text{for all open~$W \subseteq V$:} \\
    &\qquad\qquad\qquad
          \text{$\alpha_W(s|_W) = \alpha_W(t|_W)$ implies $s|_W = t|_W$} \\[0.5em]
    \Longleftrightarrow\ &
      \text{for all open~$U \subseteq X$, sections $s, t \in \F(U)$:} \\
    &\qquad
          \text{$\alpha_U(s|_U) = \alpha_U(t|_U)$ implies $s|_U = t|_U$} \\[0.5em]
    \Longleftrightarrow\ &
      \text{$\alpha$ is a monomorphism of sheaves}
  \end{align*}
}

\begin{frame}\frametitle{Building a dictionary}
  A \hil{scheme} is a locally ringed space~$(X,\O_X)$ which is
  locally isomorphic to the \hil{spectrum of a commutative ring}:
  \[ \Spec A \defeq \{ \ppp \subseteq A \,|\, \text{$\ppp$ is a prime ideal} \} \]
  \pause

  \begin{center}
    \small
    \begin{tabular}{ll}
      \toprule
      externally to $\Sh(X)$ & internally \\
      \midrule
      sheaf of rings & ring \\
      sheaf of modules & module \\
      sheaf of finite type & finitely generated module \\
      finite locally free sheaf & finite free module \\
      \bottomrule
    \end{tabular}
  \end{center}

  \hil{Motto:} \tabto{1.41cm} We can understand notions of algebraic geometry \\
  \tabto{1.41cm} as notions of linear algebra internal to~$\Sh(X)$.
\end{frame}

\begin{frame}[c]\frametitle{Using the dictionary}
  \begin{center}
    \begin{minipage}{0.75\textwidth}
      \begin{exampleblock}{}
        \justifying
        Let~$0 \to M' \to M \to M'' \to 0$ be a short exact sequence of
        modules. If~$M'$ and~$M''$ are finitely generated, so is~$M$.
      \end{exampleblock}
    \end{minipage}
    \medskip

@@

\begin{frame}[c]\frametitle{Using the dictionary}
  \begin{center}
    \visible<2->{
      \begin{minipage}{0.70\textwidth}
        \begin{exampleblock}{}
          \justifying
          Any finitely generated vector space does \emph{not not} possess a basis.
        \end{exampleblock}
      \end{minipage}
      \medskip

      \scalebox{3}{$\Downarrow$}
    }

    \begin{minipage}{0.70\textwidth}
      \begin{exampleblock}{}
        \justifying
        Any sheaf of modules of finite type on a reduced scheme is locally free
        \emph{on a dense open subset}.
      \end{exampleblock}
    \end{minipage}
  \end{center}
  \pause
  \pause

  \hil{Motto:} \tabto{1.41cm} \mbox{We can understand statements of algebraic geometry} \\
  \tabto{1.41cm} as statements of linear algebra internal to~$\Sh(X)$.
\end{frame}

\begin{frame}\frametitle{Exploring the internal world}
  Let~$X$ be a scheme. Internally to~$\Sh(X)$,
  \begin{itemize}
    \item any non-invertible element of~$\O_X$ is nilpotent,
    \item $\K_X$ is simply the total quotient ring of~$\O_X$,
    \item $\dim \O_X \leq n$ holds if and only if~$\dim X \leq n$.
  \end{itemize}

@@

\section{\texorpdfstring{The $\Diamond$-translation}{The ♦-translation}}  % XXX

@@

  Is there a \hil{free local and local-over-$B$ ring}~$A \to A'$ over~$A$?

  \[ \xymatrix{
    B \ar[r]\ar@/^1.5pc/[rrrr]\ar@/_/[rrd] & A \ar[rd] \ar[rrr] &&& R \\
    && A' \ar@{-->}[rru]
  } \]
\end{frame}
% Better decoration (local rings, local morphisms)

>>>>

\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{multicol}
\usepackage{tabto}
\usepackage{xstring}
\usepackage{mathtools}
\usepackage{soul}\setul{0.2ex}{}
\usepackage[all]{xy}
\xyoption{rotate}
\usepackage{tikz}
\usetikzlibrary{calc,shapes.callouts,shapes.arrows}

@@

\newcommand{\slogan}[1]{%
  \begin{center}%
    \setlength{\fboxrule}{0pt}%
    \setlength{\fboxsep}{-14pt}%
    {\usebeamercolor[fg]{item}\fbox{\usebeamercolor[fg]{normal
    text}\parbox{0.9\textwidth}{\begin{center}#1\end{center}}}}%
  \end{center}%
}

@@

@@

\begin{frame}\frametitle{Exploiting the internal language}
  A \hil{scheme} is a locally ringed space~$(X,\O_X)$ which is
  locally isomorphic to the \hil{spectrum of a commutative ring}:
  \[ \Spec A \defeq \{ \ppp \subseteq A \,|\, \text{$\ppp$ is a prime ideal} \} \]
  \pause
  The topos~$\Sh(X)$ is the \hil{petit Zariski topos} of~$X$.

  \begin{center}
    \small
    \begin{tabular}{ll}
      \toprule
      externally & internally to $\Sh(X)$ \\
      \midrule
      sheaf of sets & set/type \\
      morphism of sheaves & map of sets \\
      monomorphism & injective map \\
      epimorphism & surjective map \\
      sheaf of rings & ring \\
      sheaf of modules & module \\
      \bottomrule
    \end{tabular}
  \end{center}
\end{frame}

\begin{frame}\frametitle{Building a dictionary}
  \slogan{\hil{Understand notions of algebraic geometry
  as notions of algebra internal to~$\boldsymbol{\Sh(X)}$.}}
  \begin{center}
    \small
    \begin{tabular}{ll}
      \toprule
      externally & internally to $\Sh(X)$ \\
      \midrule
      sheaf of sets & set/type \\
      morphism of sheaves & map of sets \\
      monomorphism & injective map \\
      epimorphism & surjective map \\
      \midrule
      sheaf of rings & ring \\
      sheaf of modules & module \\
      sheaf of finite type & finitely generated module \\
      locally free sheaf & free module \\
      tensor product of sheaves & tensor product of modules \\
      rank function & minimal number of generators \\
      sheaf of rational functions & total quotient ring of~$\O_X$ \\
      \bottomrule
    \end{tabular}
  \end{center}

  \visible<2>{\begin{tikzpicture}[overlay]
    \draw[fill=white, draw=white, opacity=0.9] (0,0) rectangle (\paperwidth,7.1);
    \node[anchor=south west,inner sep=0] (image) at (1.8,1.3) {
      \includegraphics[width=0.7\textwidth]{images/steven-kleiman-misconceptions-about-kx}
    };
  \end{tikzpicture}}
\end{frame}

\begin{frame}[c]\frametitle{Praise for Mike Shulman}
  \centering
  \includegraphics[scale=0.4]{images/mike-shulman-stack-semantics}
  \par
\end{frame}

\begin{frame}\frametitle{Using the dictionary}
  \begin{center}
    \begin{minipage}{0.75\textwidth}
      \begin{exampleblock}{}
        \justifying
        Let~$0 \to M' \to M \to M'' \to 0$ be a short exact sequence of
        modules. If~$M'$ and~$M''$ are finitely generated, so is~$M$.
      \end{exampleblock}
    \end{minipage}
    \medskip

@@

\begin{frame}[c]\frametitle{Using the dictionary}
  \begin{center}
    \begin{minipage}{0.70\textwidth}
      \begin{exampleblock}{}
        \justifying
        Any finitely generated vector space does \emph{not not} possess a basis.
      \end{exampleblock}
    \end{minipage}
    \medskip

    \begin{minipage}{0.70\textwidth}
      \begin{exampleblock}{}
        \justifying
        Any sheaf of modules of finite type on a reduced scheme is locally free
        \emph{on a dense open subset}.
      \end{exampleblock}
    \end{minipage}
  \end{center}
\end{frame}

\begin{frame}\frametitle{A curious property}
  Let~$X$ be a scheme. Internally to~$\Sh(X)$,
  \begin{center}
    \hil{any non-invertible element of~$\boldsymbol{\O_X}$ is nilpotent.}
  \end{center}

@@

\section{\texorpdfstring{The $\Diamond$-translation}{The ♢-translation}}

@@

  Is there a \hil{free local and local-over-$\boldsymbol{B}$ ring}~$A \to A'$ over~$A$?

  \[ \xymatrix{
    B \ar[r]\ar@/^2pc/[rrrr]^{\text{local}}\ar@/_/[rrd]_[@!-33]{\text{local}} &
      A \ar[rd] \ar[rrr] &&&
      {\substack{\text{local}\\\text{\normalsize$R$}\\\phantom{\text{local}}}} \\
    && {\substack{\text{\normalsize$A'$}\\\text{local}}} \ar@{-->}[rru]_[@!35]{\text{local}}
  } \]

  \medskip
  Form limits in the category of \hil{locally ringed locales}
  by \hil{relocalizing} the corresponding limit in ringed locales.
\end{frame}

\begin{frame}
  \slogan{\hil{Understand notions and statements of algebraic geometry
  as notions and statements of algebra internal to appropriate toposes.}}

  {\vspace{-0.5em}\centering
  \includegraphics[scale=0.3]{images/external-internal-small}
  \par\medskip}

  \begin{itemize}
    \item Simplify routine proofs.
    \item Gain conceptual clarity.
    \item Develop a synthetic account of scheme theory.
  \end{itemize}
\end{frame}
% Fun facts:
% Take O_X, pullback along X_{negneg} --> X, pushforward: result is K_X.
% Very naive definition of P^n works internal to the big Zariski topos.

\backupstart

\begin{frame}[plain,c]
  \centering
  \includegraphics[scale=0.5]{images/sheafification-man}
  \sffamily

  You should totally look up:

  \hil{The Adventures of Sheafification Man}
  \par
\end{frame}

\begin{frame}\frametitle{Translating internal statements}
  Let~$X$ be a topological space (or locale) and let~$\alpha : \F \to \G$ be a
  morphism of sheaves on~$X$. Then:
  \allowdisplaybreaks
  \begin{align*}
    & \Sh(X) \models \speak{$\alpha$ is injective} \\[0.5em]
    \Longleftrightarrow\
    & \Sh(X) \models \forall s\?\F\_ \forall t\?\F\_ \alpha(s) = \alpha(t) \Rightarrow s = t \\[0.5em]
    \Longleftrightarrow\ &
      \text{for all open~$U \subseteq X$, sections $s \in \F(U)$:} \\
    &\qquad
      \text{for all open~$V \subseteq U$, sections $t \in \F(V)$:} \\
    &\qquad\qquad
        \text{for all open~$W \subseteq V$:} \\
    &\qquad\qquad\qquad
          \text{$\alpha_W(s|_W) = \alpha_W(t|_W)$ implies $s|_W = t|_W$} \\[0.5em]
    \Longleftrightarrow\ &
      \text{for all open~$U \subseteq X$, sections $s, t \in \F(U)$:} \\
    &\qquad
          \text{$\alpha_U(s|_U) = \alpha_U(t|_U)$ implies $s|_U = t|_U$} \\[0.5em]
    \Longleftrightarrow\ &
      \text{$\alpha$ is a monomorphism of sheaves}
  \end{align*}
\end{frame}

\backupend
